# Celltype annotation results

This section details results from performing celltype annotation.
Two complementary methods were used to perform celltyping:

* [`SingleR`](https://bioconductor.org/packages/release/bioc/html/SingleR.html), which uses a reference-based approach.
The [`BlueprintEncode` reference](https://rdrr.io/github/LTLA/celldex/man/BlueprintEncodeData.html) from the [`celldex` package](http://bioconductor.org/packages/release/data/experiment/html/celldex.html) was used as the reference dataset.
This reference contains stroma and immune cells identified by Blueprint and Encode.
* [`CellAssign`](https://github.com/Irrationone/cellassign), which uses a marker-gene based approach.
Marker genes for celltyping were obtained from [PanglaoDB](https://panglaodb.se/).

Cells annotated as "Unknown cell type" represent cells which could not be confidently identified using the given annotation method. 

If available, these results will be compared to celltype annotations provided by the data's originating research group. 

## Celltype annotation summary

```{r}
# Create data frame of celltypes
celltype_df <- colData(processed_sce) |>
  as.data.frame() |>
  # barcodes to a column
  tibble::rownames_to_column(var = "barcode") |>
  # `celltype_cols` was defined in the main qc_report.rmd
  dplyr::select(barcode, all_of(celltype_cols))


# Define a helper function to create tables for singler and cellassign annotations
create_celltype_n_table <- function(df, celltype_column) {
  df |>
    # create column with better name 
    dplyr::rename(`Annotated celltype` = {{celltype_column}}) |>
    # unified name for unknown cells
    dplyr::mutate(`Annotated celltype` = ifelse(
      # singler gives NA and cellassign (as we wrote it) gives "other" for unknown
      is.na(`Annotated celltype`) | `Annotated celltype` == "other", 
      "Unknown cell type",
      `Annotated celltype`)
    ) |>
    dplyr::count(`Annotated celltype`) |>
    dplyr::arrange(-n) |>
    # make this column a factor with unknown at the end
    dplyr::mutate(
      # Add percentage column
      `Percent of cells` =  paste0(round(n/sum(n) * 100, digits = 2), "%"),
      # set column order in descending order of n, but with unknown at the end
      `Annotated celltype` = forcats::fct_reorder(`Annotated celltype`, n, .desc=TRUE),
      `Annotated celltype` = forcats::fct_relevel(`Annotated celltype`, 
                                                  "Unknown cell type", 
                                                   after = Inf)
    ) |>
    # arrange on `Annotated celltype`
    dplyr::arrange(`Annotated celltype`) |>
    # set column order
    dplyr::select(
      `Annotated celltype`,
      `Number of cells` = n,
      `Percent of cells`
    ) |>
    # kable formatting
    knitr::kable(align = 'r') |>
    kableExtra::kable_styling(bootstrap_options = "striped",
                             full_width = FALSE,
                             position = "left") |>
    kableExtra::column_spec(2, monospace = TRUE)
}
```

##### `SingleR`

```{r}
create_celltype_n_table(celltype_df, singler_celltype_annotation)
```


##### `CellAssign`

```{r}
create_celltype_n_table(celltype_df, cellassign_celltype_annotation)
```


```{r, eval = has_submitter_celltypes}
asis_output("##### Submitter-provided annotations\n")
# TODO: check column name once submitter annotations are actually added to workflow
create_celltype_n_table(celltype_df, submitter_celltype_annotation)
```