# inferCNV results summary

<!--
This file is meant to be run as a child report from `main_qc_report.rmd`
-->

```{r results = 'asis', eval = has_infercnv_attempt}
# metadata(processed_sce)$infercnv_success is NA: insufficient reference cells so inferCNV wasn't run
# metadata(processed_sce)$infercnv_success is FALSE: inferCNV actually failed
if (is.na(metadata(processed_sce)$infercnv_success) &&
  sum(processed_sce$is_infercnv_reference) < params$infercnv_min_reference_cells) {
  glue::glue("
    <div class=\"alert alert-info\">
    inferCNV was not run because there were not enough normal reference cells present in the processed object.
    At least {params$infercnv_min_reference_cells} normal reference cells are required, but only {sum(processed_sce$is_infercnv_reference)} cells were present.
    </div>
  ")
} else if (!metadata(processed_sce)$infercnv_success) {
  glue::glue("
    <div class=\"alert alert-info\">
    inferCNV failed to run; no results are included in this section.
    </div>
  ")
}
knitr::knit_exit(fully = FALSE) # use fully = FALSE to ensure the main report continues knitting
```


This section provides information about the results from running `inferCNV` to quantify copy-number variation (CNV) events.
For full information about the `inferCNV` algorithm, please refer to [their official documentation](https://github.com/broadinstitute/infercnv/wiki).

`inferCNV` was run with the [`i6` HMM](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type) to quantify CNV events for each chromosome arm.
Presumed normal cells, as identified from the consensus cell types, were specified as `inferCNV`'s baseline normal reference.

## inferCNV processing information  {.tabset}

### Normal reference cells

The following table shows the consensus cell types used in the normal reference, as well as the number of cells in the processed object.
The cell types to include were chosen as those most likely to be normal cells based on this sample's diagnosis.

These cells were all specified as a single reference group for `inferCNV` to consider jointly, rather than considering each cell type separately.
Similarly, all other cells were treated as a single group of query cells.

```{r}
ref_celltypes <- metadata(processed_sce)$infercnv_reference_celltypes
colData(processed_sce) |>
  as.data.frame() |>
  dplyr::filter(consensus_celltype_annotation %in% ref_celltypes) |>
  dplyr::count(consensus_celltype_annotation) |>
  dplyr::arrange(desc(n)) |>
  dplyr::rename(
    "Consensus cell type" = consensus_celltype_annotation,
    "Number of cells" = n
  ) |>
  format_datatable()
```

### inferCNV options

`inferCNV` was run with the following options:

```{r}
# remove options that don't make sense to share in a report
metadata(processed_sce)$infercnv_options |>
  purrr::discard_at(c("out_dir", "counts_md5")) |>
  purrr::map(
    \(x) {
      if (is.null(x)) {
        "NULL"
      } else {
        # chr_exclude and min_max_counts_per_cell are vectors, so stringify for display
        paste(x, collapse = " ")
      }
    }
  ) |>
  as.data.frame() |>
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "inferCNV option",
    values_to = "value"
  ) |>
  format_datatable()
```

## inferCNV heatmap

In this section, we display the CNV heatmap exported by `inferCNV`.
The top heatmap displays expression values for all reference cells, and the bottom heatmap displays expression values for all other query cells.
Each row in the heatmap corresponds to a cell, and each column corresponds to a gene, shown ordered along chromosomes.
Note that CNV events were detected separately for each chromosome arm, as shown along the x-axis.

For more information on interpreting this heatmap, please refer to the [`inferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Interpreting-the-figure).

```{r}
knitr::include_graphics(params$infercnv_heatmap_file)
```

## UMAP colored by total CNV

This section displays the library UMAP colored by the total per-cell CNV as estimated by `inferCNV`.
This total CNV value was calculated by summing across all CNV events inferred within each cell.

```{r message=FALSE, warning=FALSE}
# create UMAP colored by infercnv_total_cnv
scater::plotUMAP(
  processed_sce,
  point_size = umap_point_size, # defined in UMAP child report
  point_alpha = 0.5,
  color_by = "infercnv_total_cnv"
) +
  scale_color_viridis_c(name = "cividis") +
  # remove axis numbers and background grid
  scale_x_continuous(labels = NULL, breaks = NULL) +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  guides(
    color = guide_colorbar(title = "Total CNV events")
  ) +
  theme_bw() +
  theme(aspect.ratio = 1)
```

## Total CNV distributions

This section displays the distributions of total CNVs across cell types.
The first ridge plot shows the CNV distributions between cell type groups: Those included in the `inferCNV` normal reference, versus all other query cells.
The second ridge plot shows the CNV distributions across consensus cell types, showing only the top seven most common cell types.

```{r fig.width = 7, fig.height = 10, warning = FALSE, message = FALSE}
# Prepare data frames for plotting
total_cnv_df <- scuttle::makePerCellDF(
  processed_sce,
  use.coldata = c("is_infercnv_reference", "consensus_celltype_annotation", "infercnv_total_cnv"),
  use.dimred = FALSE
) |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::add_count(is_infercnv_reference) |>
  dplyr::mutate(
    is_infercnv_reference = ifelse(is_infercnv_reference, "Reference cells", "Query cells"),
    is_infercnv_reference = glue::glue("{is_infercnv_reference} (N={n})")
  )

total_cnv_filtered_df <- total_cnv_df |>
  # rename for consistency with cell type report section
  dplyr::mutate(
    consensus_celltype_annotation = ifelse(
      consensus_celltype_annotation == "Unknown",
      "Unknown cell type",
      consensus_celltype_annotation
    )
  ) |>
  lump_wrap_celltypes() |>
  # rename for convenience
  dplyr::rename(celltype_lumped = consensus_celltype_annotation_lumped)

new_labels_df <- forcats::fct_count(total_cnv_filtered_df$celltype_lumped)
new_labels <- glue::glue("{new_labels_df$f} (N={new_labels_df$n})")
total_cnv_filtered_df$celltype_lumped <- factor(total_cnv_filtered_df$celltype_lumped, labels = new_labels)
total_cnv_filtered_df$celltype_lumped <- forcats::fct_rev(total_cnv_filtered_df$celltype_lumped) # reverse for display in ridge plot

# first plot: across cell type groups
cell_group_plot <- ggplot(total_cnv_df) +
  aes(x = infercnv_total_cnv, y = is_infercnv_reference, fill = is_infercnv_reference) +
  ggridges::geom_density_ridges2(scale = 0.9) +
  labs(
    x = "Total CNV events per cell",
    y = "Cell type group",
    title = "CNV between reference and query cell groups"
  ) +
  theme(legend.position = "none")

# second plot: across cell types
cell_type_plot <- ggplot(total_cnv_filtered_df) +
  aes(x = infercnv_total_cnv, y = celltype_lumped, fill = celltype_lumped) +
  ggridges::geom_density_ridges2(scale = 0.9) +
  scale_fill_brewer(palette = "Dark2", direction = -1) +
  labs(
    x = "Total CNV events per cell",
    y = "Consensus cell type",
    title = "CNV across consensus cell types"
  ) +
  theme(legend.position = "none")


cell_group_plot / cell_type_plot + plot_layout(heights = c(1, 3))
```
