# inferCNV results summary

<!--
This file is meant to be run as a child report from `main_qc_report.rmd`
-->

This section provides information about the results from running `inferCNV`.
`inferCNV` was run with the [`i6` HMM](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type) to quantify CNV events for each chromosome arm.
Presumed normal cells, as identified from the consensus cell types, were specified `inferCNV`'s normal reference baseline.

## inferCNV processing information

The following table shows the consensus cell types used in the normal reference, as well as the number of cells in the processed object.

```{r}
ref_celltypes <- metadata(processed_sce)$infercnv_reference_celltypes
colData(processed_sce) |>
  as.data.frame() |>
  dplyr::filter(consensus_celltype_annotation %in% ref_celltypes) |>
  dplyr::count(consensus_celltype_annotation) |>
  dplyr::arrange(desc(n)) |>
  dplyr::rename(
    "Consensus cell type" = consensus_celltype_annotation,
    "Number of cells" = n
  ) |>
  format_datatable()
```

`inferCNV` was run with the following options:

```{r}
# remove options that don't make sense to share in a report
metadata(processed_sce)$infercnv_options |>
  purrr::discard_at(c("out_dir", "counts_md5")) |>
  purrr::map(
    \(x) {
      if(is.null(x)) {
        "NULL"
      } else {
        # chr_exclude and min_max_counts_per_cell are vectors, so stringify for display
        paste(x, collapse = " ")
      }
    }) |>
  as.data.frame() |>
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "inferCNV option",
    values_to = "value used"
  ) |>
  format_datatable()
```


## inferCNV heatmap

In this section, we display the CNV heatmap exported by `inferCNV`.
The top heatmap displays expression values for all reference cells, and the bottom heatmap displays expression values for all other query cells.
Each row in the heatmap corresponds to a cell, and each column corresponds to a gene, shown ordered along chromosomes.
Note that CNV events were detected separately for each chromosome arm, as shown along the x-axis.

For more information on interpreting this heatmap, please refer to the [`inferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Interpreting-the-figure).

```{r}
knitr::include_graphics(params$infercnv_heatmap_file)
```
