## Dimensionality Reduction

The below plot shows the UMAP (Uniform Manifold Approximation and Projection) embeddings for each cell, coloring each cell by the total number of genes detected per cell.

```{r message=FALSE}
# create UMAP colored by number of genes detected
scater::plotUMAP(processed_sce,
  point_size = 0.3,
  point_alpha = 0.5,
  colour_by = "detected"
) +
  scale_color_viridis_c() +
  # remove axis numbers and background grid
  scale_x_continuous(labels = NULL, breaks = NULL) +
  scale_y_continuous(labels = NULL, breaks = NULL) +
  coord_fixed() +
  guides(
    color = guide_colorbar(title = "Number of \ngenes detected")
  )
```

### Expression of highly variable genes

The plots below show the same UMAP embeddings, coloring each cell by the expression level of the labeled gene.
The genes chosen for plotting are the 12 most variable genes identified in the library.
Gene symbols are used when available to label the UMAP plots.
If gene symbols are not available, the Ensembl id will be shown.

```{r message=FALSE, results='asis', fig.height = 8}
# only create plot if variable genes are present
if (!is.null(processed_meta$highly_variable_genes)) {
  # select top genes to plot
  top_genes <- processed_meta$highly_variable_genes |>
    head(n = 12)

  # grab expression for top genes from counts
  var_gene_exp <- logcounts(processed_sce[top_genes, ]) |>
    as.matrix() |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("barcode")

  # grab rowdata as data frame to later combine with gene expression data
  # rowdata contains the mapped gene symbol so can use that to label plots instead of ensembl gene id from rownames
  rowdata_df <- rowData(processed_sce) |>
    as.data.frame() |>
    tibble::rownames_to_column("ensembl_id") |>
    select(ensembl_id, gene_symbol) |>
    filter(ensembl_id %in% top_genes) |>
    mutate(
      gene_symbol = ifelse(!is.na(gene_symbol), gene_symbol, ensembl_id),
      ensembl_id = factor(ensembl_id, levels = top_genes)
    ) |>
    arrange(ensembl_id) |>
    mutate(gene_symbol = factor(gene_symbol, levels = gene_symbol))


  # extract umap embeddings as a dataframe to join with gene expression and coldata for plotting
  umap_df <- reducedDim(processed_sce, "UMAP") |>
    as.data.frame() |>
    tibble::rownames_to_column("barcode") |>
    rename(
      "UMAP1" = "V1",
      "UMAP2" = "V2"
    )

  # combine gene expression with coldata, umap embeddings, and rowdata and create data frame to use for plotting
  coldata_df <- colData(processed_sce) |>
    as.data.frame() |>
    tibble::rownames_to_column("barcode") |>
    # combine with gene expression
    left_join(var_gene_exp, by = "barcode") |>
    # combine with umap embeddings
    left_join(umap_df, by = "barcode") |>
    # combine all genes into a single column for easy faceting
    tidyr::pivot_longer(
      cols = starts_with("ENSG"),
      names_to = "ensembl_id",
      values_to = "gene_expression"
    ) |>
    # join with row data to add in gene symbols
    left_join(rowdata_df)


  ggplot(coldata_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
    geom_point(alpha = 0.1, size = 0.001) +
    facet_wrap(vars(gene_symbol)) +
    scale_color_viridis_c() +
    labs(
      color = "Log-normalized gene expression"
    ) +
    # remove axis numbers and background grid
    scale_x_continuous(labels = NULL, breaks = NULL) +
    scale_y_continuous(labels = NULL, breaks = NULL) +
    coord_fixed() +
    theme(
      legend.position = "bottom",
      axis.title = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5))
} else {
  glue::glue("
    <div class=\"alert alert-warning\">

    This library does not contain a set of highly variable genes.

    </div>
  ")
}
```
