---
params:
  merge_group: "example_group"
  merged_sce: NULL
  batch_column: "library_id"
title: "`r glue::glue('ScPCA merged report for {params$merge_group}')`"
author: "Childhood Cancer Data Lab"
date: "`r params$date`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    number_sections: false
    code_download: true
---

```{r setup, message = FALSE, echo = FALSE}
# knitr options
knitr::opts_chunk$set(
  echo = FALSE
)

library(SingleCellExperiment)
library(ggplot2)

# Set default ggplot theme, customized for UMAPs
theme_set(
  theme_bw() +
    theme(
      plot.margin = margin(rep(20, 4)),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      aspect.ratio = 1
    )
)
```

```{r}
# define a simple function for creating tables
make_kable <- function(df) {
  knitr::kable(df, align = "r") |>
    kableExtra::kable_styling(
      bootstrap_options = "striped",
      full_width = FALSE,
      position = "left"
    ) |>
    kableExtra::column_spec(2, monospace = TRUE)
}
```


```{r}
# check that input files are provided
if (is.null(params$merged_sce)) {
  stop("Must provide a merged SCE object.")
}

# define merged sce
merged_sce <- params$merged_sce

# define batch column
batch_column <- params$batch_column

# set adt/multiplex values
# this tells us if at least one library contains either of these modalities
has_adt <- "adt" %in% merged_sce$additional_modalities
multiplexed <- "cellhash" %in% merged_sce$additional_modalities
```

```{r, integration-warning, eval=FALSE}
glue::glue("
    <div class=\"alert alert-warning\">

    <b>CAUTION: The libraries included in the merged object are NOT integrated or batch-corrected.</b>

    </div>
  ")
```

```{r, ref.label=c('integration-warning'), eval=TRUE, results='asis'}
```


This report includes a brief summary of all libraries included in the merged object for `r params$merge_group`.
This merged object is intended to facilitate downstream analyses across multiple samples.

The libraries included in the merged object are _NOT integrated or batch-corrected_.
To conduct any downstream analyses that require batch correction (e.g., UMAP or clustering), batch correction must be performed separately. 

# Library information 

```{r}
# extract coldata as data frame to use to create tables and UMAPs
coldata_df <- scuttle::makePerCellDF(merged_sce, use.dimred = "UMAP") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  )

num_libraries <- merged_sce$library_id |>
  unique() |>
  length()
```

The merged object summarized in this report is comprised of `r num_libraries` individual libraries, which are described in the following table.

- `Technology version` indicates which 10X Genomics kit was used for library preparation.
- `Experimental factor ontology` indicates the associated [Experimental factor ontology](https://www.ebi.ac.uk/efo/) (EFO) for the specified 10X version. 
- `Suspension type` indicates if the library was prepared from a single-cell (`cell`) or single-nuclei (`nucleus`) suspension. 


```{r}
# table summarizing number of libraries with tech version and seq unit
tech_table <- coldata_df |>
  dplyr::select(library_id, tech_version, assay_ontology_term_id, seq_unit) |>
  dplyr::distinct() |>
  dplyr::count(tech_version, assay_ontology_term_id, seq_unit) |>
  dplyr::arrange(desc(n)) |>
  dplyr::rename(
    "Technology version" = "tech_version",
    "Experimental factor ontology" = "assay_ontology_term_id",
    "Suspension type" = "seq_unit",
    "Number of libraries" = "n"
  )

make_kable(tech_table)
```


```{r, eval=has_adt}
knitr::asis_output(glue::glue(
  "
  This merged object contains one or more libraries with accompanying antibody-derived tag (ADT) expression data.
  The following table summarizes the number of libraries with and without ADT data.
  "
))

# count number of libraries with and without ADT
adt_table <- coldata_df |>
  dplyr::select(library_id, additional_modalities) |>
  dplyr::distinct() |>
  dplyr::mutate(
    contains_adt = ifelse(additional_modalities == "adt", "Contains ADT", "No ADT")
  ) |>
  dplyr::count(contains_adt) |>
  dplyr::rename(
    "Library type" = "contains_adt",
    "Number of libraries" = "n"
  )

make_kable(adt_table)
```


```{r, eval=multiplexed }
knitr::asis_output(glue::glue(
  "
  This merged object contains one or more libraries that have been multiplexed.
  The following table summarizes the number of libraries that have and have not been multiplexed.
  "
))

# count number of libraries with and without multiplexing
multiplex_table <- coldata_df |>
  dplyr::select(library_id, additional_modalities) |>
  dplyr::distinct() |>
  dplyr::mutate(
    is_multiplexed = ifelse(additional_modalities == "cellhash", "Multiplexed", "No multiplexing")
  ) |>
  dplyr::count(is_multiplexed) |>
  dplyr::rename(
    "Library type" = "is_multiplexed",
    "Number of libraries" = "n"
  )

make_kable(multiplex_table)
```


# Sample information

```{r, results='asis'}
if (multiplexed) {
  # get the total number of libraries that have been multiplexed
  n_multiplexed <- multiplex_table |>
    dplyr::filter(`Library type` == "Multiplexed") |>
    dplyr::pull(`Number of libraries`)

  glue::glue("
    <div class=\"alert alert-info\">

    This merged object contains {n_multiplexed} libraries that have been multiplexed.
    Demultiplexing has not been performed, so sample information will not be displayed.

    </div>
  ")
}
```


```{r, eval=!multiplexed}
# get total number of samples and libraries
num_samples <- merged_sce$sample_id |>
  unique() |>
  length()
```


```{r, eval=!multiplexed, results='asis'}
glue::glue(
  "
  The merged object summarized in this report contains {num_samples} samples.
  "
)
```


```{r, eval=!multiplexed}
knitr::asis_output("
## Diagnoses

The following table shows the total number of samples associated with each diagnosis.
If a subdiagnosis is available, the total number of samples will be shown for each unique combination of diagnosis and subdiagnosis.

")
```


```{r, eval=!multiplexed}
# table of diagnosis
diagnosis_table <- coldata_df |>
  dplyr::select(sample_id, diagnosis, subdiagnosis) |>
  dplyr::distinct()

# if subdiagnosis is present, summarize both
if (!all(is.na(diagnosis_table$subdiagnosis))) {
  diagnosis_table <- diagnosis_table |>
    dplyr::count(diagnosis, subdiagnosis) |>
    dplyr::arrange(desc(n)) |>
    dplyr::rename(
      "Diagnosis" = "diagnosis",
      "Subdiagnosis" = "subdiagnosis",
      "Number of samples" = "n"
    )
} else {
  # otherwise only count diagnosis
  diagnosis_table <- diagnosis_table |>
    dplyr::count(diagnosis) |>
    dplyr::arrange(desc(n)) |>
    dplyr::rename(
      "Diagnosis" = "diagnosis",
      "Number of samples" = "n"
    )
}

make_kable(diagnosis_table)
```


```{r, eval=!multiplexed}
knitr::asis_output(
  "
## Disease stage

The following table shows both the diagnosis and at what point during disease progression (`Disease stage`) the sample was taken.

"
)
```


```{r, eval=!multiplexed}
# count number of samples per diagnosis/ disease timing combination
disease_timing_table <- coldata_df |>
  dplyr::select(sample_id, diagnosis, disease_timing) |>
  dplyr::distinct() |>
  dplyr::count(diagnosis, disease_timing) |>
  dplyr::arrange(desc(n)) |>
  dplyr::rename(
    "Diagnosis" = "diagnosis",
    "Disease stage" = "disease_timing",
    "Number of samples" = "n"
  )

make_kable(disease_timing_table)
```



## UMAPs

```{r, ref.label=c('integration-warning'), eval=TRUE, results='asis'}
```

In this section, we present UMAPs showing all cells from all libraries included in the merged object. 
For each library, a separate panel is shown, and cells from that library are colored, while all other cells are gray.

**These merged object UMAP coordinates differ from UMAPs in individual library files.**
After individual objects were merged, PCA and UMAP coordinate were recalculated so that each library would contribute equally to the overall scaling.
As a consequence, the PCA and UMAP coordinates found in the merged object will differ from the coordinates found in individual libraries.

```{r}
# Determine appropriate dimensions for UMAP based on number of libraries
n_libraries <- tech_table$`Number of libraries`

# Determine faceting layout based on n_libraries
# TODO: This needs to be revisited when we run the merged objects through
facet_ncol <- dplyr::case_when(
  n_libraries <= 20 ~ 4,
  n_libraries <= 50 ~ 6,
  n_libraries <= 80 ~ 8,
  .default = 10
)

# TODO: For now, 3 inches per UMAP row
umap_height <- ceiling(n_libraries / facet_ncol) * 3
```



```{r, eval = multiplexed, results = 'asis'}
glue::glue("
 <div class=\"alert alert-info\">
 This merged object contains multiplexed data.
 Each UMAP facet displays a given library, which may be comprised of multiple samples.
  </div>
")
```


```{r, fig.width = 8, fig.height = umap_height}
ggplot(
  coldata_df,
  aes(x = UMAP1, y = UMAP2, color = {{ batch_column }})
) +
  # set points for all "other" points
  geom_point(
    data = dplyr::select(
      coldata_df, -{{ batch_column }}
    ),
    color = "gray80",
    alpha = 0.5,
    size = 0.1
  ) +
  # set points for desired cell type
  geom_point(
    alpha = 0.5,
    size = 0.1,
    color = "firebrick3"
  ) +
  facet_wrap(
    {{ batch_column }}, # do NOT use vars()
    ncol = facet_ncol
  )
```


# Session Info
<details>
<summary>R session information</summary>
```{r session_info}
if (requireNamespace("sessioninfo", quietly = TRUE)) {
  sessioninfo::session_info()
} else {
  sessionInfo()
}
```
</details>
